RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2014-08-18 manuscripts/manuscript.Rmd
suggest phylobase (for citation at least), fix install.packages line
drop dynamic references due to CrossRef isntabilities

library("knitcitations")
cleanbib()
cite_options(citation_format = "pandoc", check.entries=FALSE)
```

@@

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data.
A common file format for sharing phylogenetic tree data has long played
an essential role in the advancement of phylogenetics research.
The rapid increase in phylogenetics data and research methods that make
use of such data has introduced many new challenges and opportunities 
to sharing this information both between software environments and 
between research groups `r citep(c("Drew PLOS Biology Tree of Life 2013", "10.1186/1756-0500-5-574"))`.

@@

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led developers to
repeatedly add small modifications to the standard to meet the needs
of their own software, which results in mutually incompatible dialects
that largely defeat the advantages of a shared  file format.
NeXML was developed to addresses each of these problems `r citep("Vos NeXML Systematic Biology")`  
by drawing on robust web standards in data representation.
The design and expressiveness of NeXML makes it much easier to meet
the rising bar for best practices in data sharing in phylogenetic 
research `r citep("10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645")`.

@@

The R language `r citep(citation())` has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations `r citep("http://cran.r-project.org/web/views/Phylogenetics.html")`,
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `ape::phylo` format provided by the
ape package `r citep(citation("ape"))`.

@@

This paper has been written as a dynamic document 
`r citep(citation("knitr")[[2]])`.  The reader may paste the code blocks
shown into their R terminal to follow along with the exmples shown.

@@

```r
install.packages("RNeXML", dependencies=c("DEPENDS", "IMPORTS", "SUGGESTS"))
```

@@

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `ape::phylo` object from the 
popular `ape` R package `r citep(citation("ape"))`. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

@@

Previously, this was possible using the `treebase` package `r citep("Boettiger Temple Lang treebase 2012")`
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard `r citep("Vos NeXML 2012")`, 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.

@@

These examples use data from the popular phylogenetics software `geiger` `r citep(citation("geiger")[[4]])`.

@@

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations", or RDFa,
as described in @Vos_2012. <!-- better to cite the W3C's RDFa TR -->
These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.

@@

Researchers cannot take advantage of NeXML metadata manipulation if
there is no metadata there.  The burden of adding good metadata remains a
significant barrier to increased sharing and reuse of scientific research
`r citep("Tenopir Data Sharing by Scientists: Practices and Perceptions")`.
`RNeXML` seeks to reduce the burden of adding
high quality, machine-readable metadata to NeXML files. This occurs on
several levels:

@@

or to papers by simply using the object's DOI (using `r citet(citation("knitcitations"))`).
For instance to add the citation information of the paper that generated the
`geospiza` phylogeny included in the `geiger` package:

@@

The `taxize_nexml` function uses the R package `taxize` 
`r citep("Chamberlain taxize F1000Research 2013")` to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see `r citet("http://www.w3.org/TR/rdf-sparql-query/")` and
`r citet("Hartig Introduction to SPARQL")`.

@@

Here we illustrate this process using the example of stochastic character
mapping `r citep("Huelsenbeck stochastic character mapping")`. A stochastic character map is simply
an annotation of the branches on a phylogeny, assigning each section
of each branch to a particular "state" (typically of a morphological
characteristic).

`r citet("Bollback simmap 2006")` provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software [^3], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

@@

`r citet(citation("phytools"))` describes the `phytools` package for R, which includes
utilities for reading, manipulating, and writing `simmap` files in R.
In this example, we also show how to define `RNeXML` functions that 
map the R representation used by Revell (an extension of the `ape` class)
into the NeXML extension we have defined by using `RNeXML` functions.

@@

We can then use various functions from `phytools` designed for `simmap`
objects `r citep(citation("phytools"))`, such as the plotting function:

@@

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity (e.g. `r citet("Whitlock data archiving Evolution 2010")`).  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier (@Tenopir_2011, `r citet("10.2139/ssrn.1550193")`),
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [@Drew_2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.

@@

There are many promising directions for further development of the
`RNeXML` package.  The project's [Milestones page](https://github.com/ropensci/RNeXML/issues/milestones) page provides a list of
steps currently planned or under consideration.  In particular, these
milestones include increased support for interactive mode of entering data
(wizards), conversions to and from other existing R phylogenetics formats
such those found in the `phylobase` package `r citep(citation("phylobase"))`,
and support for integration of semantics defined through the
Web Ontology language (OWL), as discussed in the Challenges section.

>>>>

```

@@

Though R provides the most extensive and widely used collection of
comparative phylogenetic methods tools, R users lack a convenient,
extensible format to exchange and archive comparative methods data.
A common file format for sharing phylogenetic tree data has long played
an essential role in the advancement of phylogenetics research.
The rapid increase in phylogenetics data and research methods that make
use of such data has introduced many new challenges and opportunities 
to sharing this information both between software environments and 
between research groups [@Drew_2013; @Stoltzfus_2012].

@@

For example, the currently popular NEXUS file format cannot consistently
represent horizontal gene transfer or ambiguity in reading a character
(such as a DNA sequence base pair).  This has led developers to
repeatedly add small modifications to the standard to meet the needs
of their own software, which results in mutually incompatible dialects
that largely defeat the advantages of a shared  file format.
NeXML was developed to addresses each of these problems [@Vos_2012]  
by drawing on robust web standards in data representation.
The design and expressiveness of NeXML makes it much easier to meet
the rising bar for best practices in data sharing in phylogenetic 
research [@Cranston_2014].

@@

The R language [@R] has become an important environment for 
phylogenetics research, with over 30 packages available including several
with over 1000 citations [@taskview],
though stand-alone programs remain an important part of this ecosystem
for tasks such as generating phylogenetic trees from sequence data.  Within 
this ecosystem a variety of different data formats for
phylogenetic data have been developed and modified. Most but not all 
are based on the simple and popular `ape::phylo` format provided by the
ape package [@Paradis_2004].

@@

This paper has been written as a dynamic document 
[@Xiei_2013].  The reader may paste the code blocks
shown into their R terminal to follow along with the exmples shown.

@@

```r
install.packages("RNeXML", dependencies=TRUE)
```

@@

The resulting `nex` object is an R object representation of the NeXML. This is 
a native R object, of the `nexml` class, for representing phylogenetic trees (as well as 
character traits, and metadata), analogous to the `ape::phylo` object from the 
popular `ape` R package [@Paradis_2004]. While R functions and packages 
could be written to use this `nexml` object directly, most users will prefer to use
already existing methods provided by packages they know. From a `nexml` object, we can 
therefore extract any phylogenies it contains in the `ape::phylo` format. In this 
case we will get back a `multiPhylo` object, since the file contains two phylogenies.

@@

Previously, this was possible using the `treebase` package [@Boettiger_2012"]
for R, which relied on the NEXUS parser provided in the `ape` package.
Due to inconsistencies present in the NEXUS standard [@Vos_2012], 
a significant number <!-- State explicitly --> of TreeBASE NEXUS
files could not be read in this way.  Using the more precisely defined
NeXML standard, every valid NeXML document on TreeBASE can now successfully
be parsed to R objects.

@@

These examples use data from the popular phylogenetics software `geiger` [@Harmon_2008].

@@

At the heart of the NeXML standard is the ability to encode arbitrary
metadata in precisely specified, machine-readable formats that maintain
backwards compatibility with any existing NeXML parser. NeXML uses the
"Resource Description Framework in Annotations", or RDFa,
as described in @Vos_2012 or @W3C_2014. 
These annotations can be used to add 
information to any particular element of the data, (a tree, node, edge,
or taxonomic unit; a character matrix or specific trait, or even another 
metadata annotation or the NeXML data file as a whole). `RNeXML` provides
a range of tools for accessing metadata embedded in a NeXML file.

@@

Researchers cannot take advantage of NeXML metadata manipulation if
there is no metadata there.  The burden of adding good metadata remains a
significant barrier to increased sharing and reuse of scientific research
[@Tenopir_2011].
`RNeXML` seeks to reduce the burden of adding
high quality, machine-readable metadata to NeXML files. This occurs on
several levels:

@@

or to papers by simply using the object's DOI (using @Boettiger_2014).
For instance to add the citation information of the paper that generated the
`geospiza` phylogeny included in the `geiger` package:

@@

The `taxize_nexml` function uses the R package `taxize` 
[@Chamberlain_2013] to check each taxon label against the NCBI database.
If a unique match is found, a metadata annotation is added to the taxon
providing the NCBI identification number to the taxonomic unit.  If no
match is found, the user is warned to check for possible typographic
errors in the taxonomic labels provided.

@@

The most powerful way to take advantage of the machine-readable nature
of the metadata involves the construction of SPARQL queries. This
approach can exploit the reasoning and logical deduction capacity of
formal vocabularies (ontologies). For a proper introduction to SPARQL
queries, see @W3C_2014 or @Hartig_2012).

@@

Here we illustrate this process using the example of stochastic character
mapping [@Huelsenbecki_2003]. A stochastic character map is simply
an annotation of the branches on a phylogeny, assigning each section
of each branch to a particular "state" (typically of a morphological
characteristic).

@Bollback_2006 provides a widely used stand-alone software implementation
of this method in the software `simmap`, which modified the standard
Newick tree format to express this additional information. This can
break compatibility with other software [^3], and creates a format that
cannot be interpreted without additional information describing this
convention.  By contrast, the NeXML extension is not only backwards
compatible but contains a precise and machine-readable description of
what it is encoding.

@@

@Revell_2012 describes the `phytools` package for R, which includes
utilities for reading, manipulating, and writing `simmap` files in R.
In this example, we also show how to define `RNeXML` functions that 
map the R representation used by Revell (an extension of the `ape` class)
into the NeXML extension we have defined by using `RNeXML` functions.

@@

We can then use various functions from `phytools` designed for `simmap`
objects [@Revell_2012], such as the plotting function:

@@

Data archiving is increasingly required by journals in evolutionary biology and 
biodiversity (e.g. @Whitlock_2010).  The burden of preparing and submitting properly
annotated data to archives continues to be a significant barrier [@Tenopir_2011; @Stodden_2014],
and many phylogenetic trees upon which studies are based are inaccessible or lost 
to the research community [@Drew_2013].  `RNeXML` seeks to lower these barriers
by providing functions to immediately archive NeXML files in scientific repositories.

@@

There are many promising directions for further development of the
`RNeXML` package.  The project's [Milestones page](https://github.com/ropensci/RNeXML/issues/milestones) page provides a list of
steps currently planned or under consideration.  In particular, these
milestones include increased support for interactive mode of entering data
(wizards), conversions to and from other existing R phylogenetics formats
such those found in the `phylobase` package [@phylobase],
and support for integration of semantics defined through the
Web Ontology language (OWL), as discussed in the Challenges section.
